---
title: "Flow structures: if-else and loops. Functions in R"
subtitle: "Practical workbooks of Data Programming in Master in Computational Social Science  (2024-2025)"
author: "Javier Álvarez Liébana"
format:
  html:
    theme: [default, style.scss]
    toc: true
    toc-title: Índice
    toc-depth: 5
    toc-location: left
    number-sections: true
embed-resources: true
execute: 
  echo: true
---

```{r}
#| echo: false
setwd(dir = getwd())
```

## Flow structures

A [**flow or control structure**]{.hl-yellow} consists of a series of commands oriented to [**decide the path**]{.hl-yellow} that your code must follow

* If condition A is met, what happens?

* What if B happens?

* How can I repeat the same expression (depending on a variable)?


If you have programmed before, you may be familiar with what are known as [**conditional structures**]{.hl-yellow} such as `if (bla bla) {...} else {...}` or [**loops**]{.hl-yellow} `for/while` (to be avoided whenever possible).


### Conditional structures

#### If

One of the most famous control structures are those known as [**conditional structures**]{.hl-yellow} `if`.

> IF a set of conditions is met (TRUE), then execute whatever is inside the curly brackets.


For example, the structure `if (x == 1) { code A }` what it will do is [**execute code A in braces**]{.hl-yellow} but [**ONLY IF**]{.hl-purple} the [**condition in brackets is true**]{.hl-purple} (only if `x` is 1). In any other case, it will do nothing

Let's define a vector of ages of 8 people


```{r}
ages <- c(14, 17, 24, 56, 31, 20, 87, 73)
ages < 18
```



Our conditional structure will do the following: [**if there is a minor, it will print**]{.hl-yellow} a message.

```{r}
if (any(ages < 18)) { 
  
  print("There is a minor")
  
}
```



```{r}
#| eval: false
if (any(ages < 18)) { 
  
  print("There is a minor")
  
}
```


In case the [**conditions are not true**]{.hl-yellow} inside `if()` (`FALSE`), nothing happens.


```{r}
if (all(ages >= 18)) { 
  
  print("All of them are of legal age")
  
}
```

**We get no message** because the condition `all(ages >= 18)` is not `TRUE`, so it does not execute anything.


#### If-else

The structure `if (condition) { code A }` can be combined with an `else { code B }`: when the [**condition is not checked**]{.hl-yellow}, it will [**execute the alternative code B**]{. hl-yellow} inside `else { }`, allowing us to decide what happens when it is satisfied and when it is not

For example, `if (x == 1) { code A } else { code B }` will execute A if `x` is equal to 1 and B in any other case.

```{r}
if (all(ages >= 18)) { 
  
  print("All of them are of legal age")
  
} else {
  
  print("There is a minor")
}
```



Esta estructura `if - else` puede ser [**anidada**]{.hl-yellow}: imagina que queremos ejecutar un código si todos son menores; si no sucede, pero todos son mayores de 16, hacer otra cosa; en cualquier otra cosa, otra acción.

```{r}
if (all(ages >= 18)) { 
  
  print("All of them are of legal age")
  
} else if (all(ages >= 16)) {
  
  print("There is a minor but all of them are greater or equal to 16 years old")
  
} else { print("There are any persons under 16 years of age") }
```

::: callout-note
## Tip

You can **collapse the structures** by clicking on the left arrow in your script.

:::


#### If-else vectorized


This conditional structure can be [**vectorized**]{.hl-yellow} (in a single line) with `if_else()` (from the `{dplyr}` package), whose arguments are

* the condition to evaluate

* what happens when it is met and when not

* an optional argument for when the condition to evaluate is `NA`

We will label without are greater/lesser and an `unknown` when we don't know.

```{r}
library(dplyr)
ages <- c(NA, ages)
if_else(ages >= 18, "legal age", "minor", missing = "unknown")
```

In `R` base there is `ifelse()`: it does not let you specify what to do with the absent ones but allows you to specify different types of data in `TRUE` and `FALSE`.

### 💻 It's your turn


[**Try to perform the following exercises without looking at the solutions**]{style="color:#444442;"}

::: panel-tabset

### [**Exercise 1**]{.hl-yellow}

📝 What will be the output of the following code?

```{r}
#| eval: false
if_else(sqrt(9) < 2, sqrt(9), 0)
```

```{r}
#| eval: false
#| code-fold: true

The output is 0 since sqrt(9) equals 3, and since it is not less than 2, it returns the second argument which is 0.
```

### [**Exercise 2**]{.hl-yellow}

📝 What will be the output of the following code?

```{r}
#| eval: false
x <- c(1, NA, -1, 9)
if_else(sqrt(x) < 2, 0, 1)
```

```{r}
#| eval: false
#| code-fold: true

The output is the vector c(0, NA, NA, 1) since sqrt(1) is less than 2, sqrt(9) is not, and in the case of both sqrt(NA) (root of absent) and sqrt(-1) (returns NaN, not a number), its square root cannot be checked whether it is less than 2 or not, so the output is NA.
```


### [**Exercise 3**]{.hl-yellow}

📝 Modify the code below so that, when the square root of a number cannot be verified to be less than 2, it returns -1.

```{r}
#| eval: false
x <- c(1, NA, -1, 9)
if_else(sqrt(x) < 2, 0, 1)
```

```{r}
#| eval: false
#| code-fold: true
x <- c(1, NA, -1, 9)
if_else(sqrt(x) < 2, 0, 1, missing = -1)
```


### [**Exercise 4**]{.hl-yellow}

📝 What are the values of `x` and `y` of the lower code for `z <- 1`, `z <- -1` and `z <- -5`?

```{r}
#| eval: false
z <- -1
if (z > 0) {
  
  x <- z^3
  y <- -sqrt(z)
  
} else if (abs(z) < 2) {
  
  x <- z^4
  y <- sqrt(-z)
  
} else {
  
  x <- z/2
  y <- abs(z)
  
}
```

```{r}
#| eval: false
#| code-fold: true
In the first case x = 1 and y = -1. In the second case x = 1 and y = 1. In the third case -1 and 2.
```


### [**Exercise 5**]{.hl-yellow}

📝 What will happen if we execute the code below?

```{r}
#| eval: false
z <- "a"
if (z > 0) {
  
  x <- z^3
  y <- -sqrt(z)
  
} else if (abs(z) < 2) {
  
  x <- z^4
  y <- sqrt(-z)
  
} else {
  
  x <- z/2
  y <- abs(z)
  
}
```

```{r}
#| code-fold: true
#| eval: false

# will give error since it is not a numeric argument
Error in z^3 : non-numeric argument to binary operator
```


### [**Exercise 6**]{.hl-yellow}

📝 From the `{lubridate}` package, the `hour()` function returns the time of a given date, and the `now()` function returns the date and time of the current time. With both functions, have `cat()` (`cat()`) print “good night” only after 21:00.

```{r}
#| code-fold: true
#| eval: false

# loading library
library(lubridate)

# Current date-time
current_dt <- now()

# If structure
if (hour(current_dt) > 21) {
  
  cat("Good night") # print or cat (two ways of printing)
}
```


:::

### Loops

Although in most occasions they can be replaced by other more efficient and readable structures, it is important to know one of the most famous control expressions: the [**loops**]{.hl-yellow}.

* `for { }`: allows [**repeating the same code**]{. hl-yellow} in a [**prefixed and known number**]{.hl-purple} of times.

* `while { }`: allows [**repeating the same code**]{.hl-yellow} but in an [**undetermined number of times**]{.hl-purple} (until a **condition** is no longer fulfilled).


#### For loop 

A [**for**]{.hl-yellow} loop is a structure that allows to [**repeat**]{.hl-yellow} a set of commands a [**finite, prefixed and known number of times**]{.hl-purple} given a set of indices.

Let's define a vector `x <- c(0, -7, 1, 4)` and another empty variable `y`. After that we will define a for loop with `for () { }`: inside the brackets we will indicate an index and some values to traverse, inside the braces the code to execute in each iteration (in this case, fill `y` as `x + 1`).


```{r}
x <- c(0, -7, 1, 4)
y <- c()

for (i in 1:4) {
  
  y[i] <- x[i] + 1
  
}
```

Note that because `R` works in a [**default vector**]{.hl-yellow} manner, the loop is the same as doing `x + 1` directly.

```{r}
x <- c(0, -7, 1, 4)
y <- c()

for (i in 1:4) {
  
  y[i] <- x[i] + 1
  
}
y

y2 <- x + 1
y2
```

Another common option is to indicate the indexes “automatically”: from the first `1` to the last (corresponding to the length of x `length(x)`).

```{r}
x <- c(0, -7, 1, 4)
y <- c()

for (i in 1:length(x)) {
  
  y[i] <- x[i] + 1
  
}
y
```


Thus the [**general structure of a for-loop**]{.hl-yellow} will always be as follows

```{r}
#| eval: false
for (index in set) { 
  
  código (usually depending on index)
  
}
```

In the case of for loops [**ALWAYS**]{.hl-green} we know how many iterations we have (as many as there are elements in the set to be indexed). We can see another example of a [**combining numbers and text**]{.hl-yellow} loop: we define a vector of ages and names, and print the i-th name and age.

```{r}
library(glue)
names <- c("Javi", "Sandra", "Carlos", "Marcos", "Marta")
ages <- c(33, 27, 18, 43, 29)

for (i in 1:5) { 
  
  print(glue("{names[i]} are {ages[i]} old")) 
  
}
```

Although they are usually indexed with numeric vectors, loops can be [**indexed on any vector structure**]{.hl-yellow}, regardless of the type of the set.

```{r}
library(stringr)
week_days <- c("monday", "tuesday", "wednesday", "thursday",
               "friday", "saturday", "sunday")

for (days in week_days) {
  
  print(days)
}
```

##### for + if-else

Let's **combine conditional structures and loops**: using the `swiss` set of the `{datasets}` package, let's assign `NA` if the fertility values are greater than 80.

```{r}
for (i in 1:nrow(swiss)) {
  
  if (swiss$Fertility[i] > 80) { 
    
    swiss$Fertility[i] <- NA
    
  }
}
```


This is «the same» as a vectorized `if_else()`.

```{r}
data("swiss")
swiss$Fertility <- if_else(swiss$Fertility > 80, NA, swiss$Fertility)
```


#### While loop

Another way to create a loop is with the `while { }` structure, which will loop [**an unknown number of times**]{.hl-yellow}, until a condition [**stops being met**]{.hl-yellow} (in fact it may never end). For example, we will inialize a variable `times <- 1`, which we will increment at each step, and we will not exit the loop until `times > 3`.

```{r}
times <- 1
while(times <= 3) {
  
  print(glue("Not yet, we are in the {times}-th iteration")) 
  times <- times + 1
  
}
print(glue("Now! We are in the {times}-th iteration")) 
```

A `while` loop will always look like this

```{r}
#| eval: false
while(condition) {
  
  code to be executed while condition is TRUE
  # usually some variable is updated here
  
}
```


What happens when the [**condition is never FALSE**]{.hl-yellow}? Try it yourself

```{r}
#| eval: false
while (1 > 0) {
  
  print("Press ESC to exit")
  
}
```

&nbsp;

::: callout-warning
## Warning

A `while { }` loop can be quite “dangerous” if we do not control well how to stop it.

:::

##### break and next

We have two reserved commands to [**abort a loop or force it forward**]{.hl-yellow}:

* `break`: allows [**abort a loop**]{.hl-yellow} even if its end has not been reached

```{r}
for(i in 1:10) {
  if (i == 3) {
    
    break # if i = 3, we abort
    
  }
  print(i)
}
```


* `next`: [**forces a loop to advance to the next iteration**]{.hl-yellow} 

```{r}
for(i in 1:5) {
  if (i == 3) {
    
    next # if i = 3, we advance to the next iteration
    
  }
  print(i)
}
```


### 💻 It's your turn

[**Try to perform the following exercises without looking at the solutions**]{style="color:#444442;"}


::: panel-tabset

### [**Exercise 1**]{.hl-yellow}

📝 Modify the code below to print a message on the screen if and only if all the data in `airquality` is for a month other than January.

```{r}
#| eval: false
library(datasets)
months <- airquality$Month

if (months == 2) {
  print("No data in January")
}
```

```{r}
#| code-fold: true
#| eval: false
library(datasets)
months <- airquality$Month

if (all(months != 1)) {
  print("No data in January")
}
```


### [**Exercise 2**]{.hl-yellow}

📝 Modify the code below to store in a variable called `temp_high` a `TRUE` if any of the records has a temperature above 90 degrees Fahrenheit and `FALSE` in any other case.

 
```{r}
#| eval: false
temp <- airquality$Temp

if (temp == 100) {
  print("Some of the records have temperatures in excess of 90 degrees Fahrenheit")
}
```

 
```{r}
#| eval: false
#| code-fold: true
# Option 1
temp <- airquality$Temp
temp_high <- FALSE
if (any(temp > 90)) {
   temp_high <- TRUE
}

# Option 2
temp_high <- any(airquality$Temp > 90)
```

### [**Exercise 3**]{.hl-yellow}

📝 Modify the code below to design a `for` loop of 5 iterations that only loops through the first 5 odd (and at each step of the loop prints them)

```{r}
#| eval: false
for (i in 1:5) {
  
  print(i)
}
```

```{r}
#| eval: false
#| code-fold: true
for (i in c(1, 3, 5, 7, 9)) {
  
  print(i)
}
```

### [**Exercise 4**]{.hl-yellow}

📝 Modify the code below to design a `while` loop that starts with a counter `count <- 1` and stops when it reaches 6

```{r}
#| eval: false
count <- 1
while (count == 2) {
  
  print(count)
}
```

```{r}
#| eval: false
#| code-fold: true
count <- 1
while (count < 6) {
  
  print(count)
  count <- count + 1
  
}
```


:::



## 🐣 Case study I: simulation study


To practice control structures we are going to perform a [**simulation exercise**]{.hl-yellow}

### Question 1


> Define a variable called `amount` that starts at 100. Design a loop of 20 iterations where on each iteration, amount is reduced to half its value. Think about what kind of loop structure you should use. The final value of `amount` should be `0.000095367` (approx).

```{r}
#| code-fold: true
#| eval: false

# We use a for since we know the number of iterations
# by default (and it does not depend on anything).

# we initially define amount in 100
amount <- 100 

# for the loop we use e.g. i as index, ranging from 1 to 20
for (i in 1:20) {
  
  # the code is the same and does not depend on i
  amount <- amount/2
}
amount
```


### Question 2

> Design a loop structure so that you find the iteration where `amount` is less than 0.001 for the first time. Once found save it in `iter` and stop the loop.


```{r}
#| code-fold: true
#| eval: false

# two ways: for and while

# for
amount <- 100 

# we already know that in 20 it is less than 0.001 so we can set
# that amount as a ceiling knowing that it will not be reached
for (i in 1:20) {
  
  # if it is still not less, we continue dividing
  if (amount >= 0.001) {
    
    amount <- amount/2
    
  } else {
    
    # if it is already smaller, we save the iteration (think why i - 1)
    iter <- i - 1 
    
    # and abort it
    break
  }
  
}

# while
amount <- 100 

iter <- 0 # we must initialize the iterations

# we don't know how many iterations, only that it should stop when
# amount is below that amount
while (amount >= 0.001) {
  
  amount <- amount/2
    
  # classic while structure: if iteration runs
  # we update a value (in this case to count one iteration)
  iter <- iter + 1
}

iter
```


### Question 3

> In `R` we have the function `%%`: if we put `a %% b` it returns the remainder that would give the division $a/b$. For example, `4 %%% 2` gives 0 since 4 is an even number (that is, its remainder when dividing by 2 is 0). If we put `13 %% 5` we get 3, since the remainder of dividing 13 by 5 is 3.

```{r}
# Remainder by dividing by 2
3 %% 2
4 %% 2
5 %% 2
6 %% 2

# Remainder by dividing by 3
9 %% 3
10 %% 3
11 %% 3
12 %% 3
```

> Starting at an initial amount `initial_amount` of 100 (euros), design a loop that adds 3€ plus the iteration you are on if the current amount is even and subtracts 5€ minus the iteration you are on if it is odd, UNLESS the amount is already equal or below 0 (in that case it should neither add nor subtract). Example: if amount is 50 euros and you are in iteration 13, it will add 3 + 13 (66 in total); if amount is 51 euros and you are in iteration 13, it will subtract 5 + 13 (33 in total); if amount is -2 euros and you are in iteration 13, it will add 3 + 13 (14 in total); if amount is -1 euros and you are in iteration 13, it will do nothing. Save the resulting amounts for each iteration (maximum of 150 iterations). Start from iteration 2

```{r}
#| code-fold: true
#| eval: false

initial_amount <- 100
amount <- c(initial_amount, rep(NA, 149))
for (i in 2:150) {
  
  if (amount[i - 1] %% 2 == 0) {
    
    amount[i] <- amount[i - 1] + 3 + i
    
  } else if (amount[i - 1] > 0) {
    
    amount[i] <- amount[i - 1] - (5 + i)
    
  } else {
    
    amount[i] <- amount[i - 1]
    
  }
}
```

> What happened?


## Functions in R


Not only can we use **default functions** that come already loaded in packages, we can also [**create our own functions**]{.hl-yellow} to **automate tasks**. How to [**create our own function**]{.hl-purple}?

### Basic scheme

 Let's look at its **basic scheme**:
 
* [**Name**]{.hl-yellow}: for example `name_fun` (no spaces or strange characters). To the name we [**assign the reserved word**]{.hl-yellow} `function()`.

* Define [**input arguments**]{.hl-yellow} (inside `function()`).

* [**Body**]{.hl-yellow} of the function inside `{ }`.

* We end the function with the [**output arguments**]{.hl-yellow} with `return()`.

```{r}
#| eval: false
name_fun <- function(arg1, arg2, ...) {
  
  code to be executed
  
  return(var_output)
  
}
```


* `arg1, arg2, ...`: will be the [**input arguments**]{.hl-yellow}, the arguments that the function takes to execute the code inside.

* `code`: lines of code that we want to [**execute the function**]{.hl-yellow}.

* `return(var_output)`: the [**output arguments**]{.hl-yellow} will be entered.


```{r}
#| eval: false
name_fun <- function(arg1, arg2, ...) {
  
  # Code to be executed
  code
  
  # Output
  return(var_output)
  
}
```

::: callout-important
## Important

All variables that we define inside the function are [**LOCAL variables: they will only exist inside the function**]{.hl-yellow} unless we specify otherwise.

:::


Let's look at a very simple example of a function for [**calculating the area of a rectangle**]{.hl-yellow}.

Since the area of a rectangle is calculated as the **product of its sides**, we will need just that, its sides: those will be the [**input arguments**]{.hl-yellow} and the [**value to return**]{.hl-purple} will be just its **area** ($side_1 * side_2$).

```{r}
# We define the name of function and input arguments
compute_area <- function(side_1, side_2) {
  
  area <- side_1 * side_2
  return(area)
  
}
```


We can also make a direct definition of variables **without storing along the way**.

```{r}
# We define the name of function and input arguments
compute_area <- function(side_1, side_2) {
  
  return(side_1 * side_2)
  
}
```

&nbsp;

[**How to apply our function?**]{.hl-yellow}

```{r}
compute_area(5, 3) # area of 5 x 3 rectangle
compute_area(1, 5) # area of 1 x 5 rectangle
```


::: callout-tip

Although it is not necessary, it is [**recommendable to make explicit the calling of the arguments**]{.hl-green}, specifying in the code what value is for each argument so that it does not depend on its order, making the code more readable.

:::


```{r}
compute_area(side_1 = 5, side_2 = 3) # area of 5 x 3 rectangle
compute_area(side_2 = 3, side_1 = 5) # area of 5 x 3 rectangle
```


### Default arguments

Imagine now that we realize that 90% of the time we use such a function to [**default calculate the area of a square**]{.hl-yellow} (i.e., we only need one side). To do this, we can define [**default arguments**]{.hl-yellow} in the function: they will take that value unless we assign another one.

Why not assign `side_2 = side_1` **default**, to save lines of code and time?


```{r}
compute_area <- function(side_1, side_2 = side_1) {
  
  # Code to be executed
  area <- side_1 * side_2
  
  # Output
  return(area)
  
}
```


Now [**default**]{.hl-yellow} the second side will be equal to the first (if added it will use both).


```{r}
compute_area(side_1 = 5) # square
compute_area(side_1 = 5, side_2 = 7) # rectangle
```


### Multiple outputs

Let's complicate the function a bit and add in the output the values of each side, labeled `side_1` and `side_2`, [**packing the output in a vector**]{.hl-yellow}.

```{r}
#| code-line-numbers: "7-8"
compute_area <- function(side_1, side_2 = side_1) {
  
  # Code
  area <- side_1 * side_2
  
  # Output
  return(c("area" = area, "side_1" = side_1, "side_2" = side_2))
  
}
```



We can complicate the output a little more by adding a fourth variable that tells us, depending on the arguments, [**whether rectangle or square**]{.hl-yellow}, having to add a character (or logic) variable in the output.


```{r}
#| code-line-numbers: "7-9"
compute_area <- function(side_1, side_2 = side_1) {
  
  # Code
  area <- side_1 * side_2
  
  # Output
  return(c("area" = area, "side_1" = side_1, "side_2" = side_2,
           "type" = if_else(side_1 == side_2, "square", "rectangle")))
  
}
compute_area(5, 3)
```


[**Problem**]{.hl-red}: when trying to put numbers and text together, it converts everything to numbers. We could store it all in a `tibble()` as we have learned or in an object known in `R` as [**lists**]{.hl-yellow} (we will see it later).


### Order of arguments

Before we did not care about the order of the arguments, but now the [**order of the input arguments matters**]{.hl-yellow}, since we include `side_1` and `side_2` in the output. 



::: callout-note
## Tip

As mentioned, it is highly recommended to make the function call [**explicitly setting the arguments**]{.hl-yellow} to improve **legibility and interpretability**.

```{r}
# Equivalent to compute_area(5, 3)
compute_area(side_1 = 5, side_2 = 3)
```

:::


It seems silly what we have done but we have crossed an important frontier: we have gone from [**consuming knowledge**]{.hl-yellow} (code from other packages, elaborated by others), to [**generating knowledge**]{.hl-purple}, creating our own functions.

Functions are going to be key in your day-to-day work because they will allow you to automate code that you are going to repeat over and over again: by packaging that code under an alias (function name) you will [**be able to use it over and over again without programming it (so doing twice as much work will not imply working twice as much)**]{.hl-yellow}


### Local vs global variables

An important aspect to think about with functions: what happens if we [**name a variable inside**]{.hl-yellow} a function to which we have **forgotten to assign** a value inside the function.

We must be cautious when using functions in `R`, since due to the [**"lexicographic rule ”**]{.hl-yellow}, if a variable is not defined inside the function, `R` will [**look for that variable in the environment**]{.hl-purple} of variables.


```{r}
x <- 1
fun_example <- function() {
    
  print(x) # No output, just doing an action
}
fun_example()
```


If a variable [**is already defined outside the function (global environment)**]{.hl-yellow}, and is also used inside changing its value, the value [**only changes inside**]{.hl-yellow} but [**not in the global environment**]{.hl-red}.

```{r}
x <- 1
fun_example <- function() {
    
  x <- 2
  print(x) # value inside of function
}
```

```{r}
# value inside of function (local)
fun_example()
# value output of function (global)
print(x)
```



If we want it to change locally as well as [**globally**]{.hl-yellow} we must use the [**double assignment**]{.hl-yellow} (`<<-`).

```{r}
x <- 1
y <- 2
fun_example <- function() {
  
  # no change in a global way, just locally
  x <- 3 
  # change in a global way
  y <<- 0 #<<
  
  print(x)
  print(y)
}

fun_example() # value inside function (local)
x # global value
y # global value
```

### 💻 It's your turn 

[**Try to perform the following exercises without looking at the solutions**]{style="color:#444442;"}

::: panel-tabset

### [**Ex 1**]{.hl-yellow}

📝 Modify the code below to define a function called `sum_function`, so that given two elements, it returns their sum.

```{r}
#| eval: false
name <- function(x, y) {
  sum_output <- # code
  return()
}
# we apply the function
sum_function(3, 7)
```

```{r}
#| code-fold: true
#| eval: false
sum_function<- function(x, y) {
  sum_output <- x + y
  return(sum_output)
}
sum_function(3, 7)
```

### [**Ex 2**]{.hl-yellow}

📝 Modify the code below to define a function called `product_function`, so that given two elements, it returns their product, but by default it calculates the square

```{r}
#| eval: false
name <- function(x, y) {
  prod_output <- # code
  return()
}
product_function(3)
product_function(3, -7)
```

```{r}
#| code-fold: true
#| eval: false
product_function <- function(x, y = x) {
  
  prod_output <- x * y
  return(prod_output)
  
}
product_function(3)
product_function(3, -7)
```

### [**Ex 3**]{.hl-yellow}

📝 Define a function called `equal_names` that, given two names, tells us if they are equal or not. Do this by considering case-sensitive, and case-insensitive. Use the `{stringr}` package.

```{r}
#| code-fold: true
#| eval: false
# Case-sensitive
equal_names <- function(person_1, person_2) {
  
  return(person_1 == person_2)
  
}
equal_names("Javi", "javi")
equal_names("Javi", "Lucía")

# Case-insensitive
equal_names <- function(person_1, person_2) {
  
  return(toupper(person_1) == toupper(person_2))
  
}
equal_names("Javi", "javi")
equal_names("Javi", "Lucía")
```

### [**Ex 4**]{.hl-yellow}

📝 Create a function called `compute_BMI` that, given two arguments (weight and height in meters) and a name, returns a list with the BMI ($weight/(height^2)$) and the name.

```{r}
#| code-fold: true
#| eval: false
compute_BMI <- function(name, weight, height) {
  
  return(list("name" = name, "BMI" = weight/(height^2)))
  
}
```

### [**Ex 5**]{.hl-yellow}

📝 Repeat the previous exercise but with another optional argument called units (by default, `units = “meters”`). Develop the function so that it does the right thing if `units = “meters”` and if `units = “centimeters”`.

```{r}
#| code-fold: true
#| eval: false
compute_BMI <- function(name, weight, height, units = "meters") {
  
  return(list("name" = name,
              "BMI" = weight / (if_else(units == "meters", height, height/100)^2)))
  
}
```

 
### [**Ex 6**]{.hl-yellow}

📝 Create a fictitious tibble of 7 persons, with three variables (invent name, and simulate weight, height in centimeters), and apply the defined function so that we obtain a fourth column with their BMI.

```{r}
#| code-fold: true
#| eval: false
data <-
  tibble("name" = c("javi", "sandra", "laura",
                       "ana", "carlos", "leo", NA),
         "weight" = rnorm(n = 7, mean = 70, sd = 1),
         "height" = rnorm(n = 7, mean = 168, sd = 5))

data |> 
  mutate(BMI = compute_BMI(name, weight, height, units = "centimeters")$BMI)
```


### [**Ex 7**]{.hl-yellow}

📝 Create a function called `shortcut` that has two numeric arguments `x` and `y`. If both are equal, you should return `equal` and have the function terminate automatically (think about when a function exits). WARNING: `x` and `y` could be vectors. If they are different (of equal length) calculate the proportion of different elements. If they are different (being of different length), it returns the elements that are not common.

```{r}
#| code-fold: true
#| eval: false
shortcut <- function(x, y) {
  
  if (all(x == y) & length(x) == length(y)) { return("equal") }
  else {
   
    if (length(x) == length(y)) {
      
      n_diff <- sum(x != y) / length(x)
      return(n_diff)
      
    } else {
      
      diff_elem <- unique(c(setdiff(x, y), setdiff(y, x)))
      return(diff_elem)
    }
    
  }
}
```


:::

## 🐣 Case study II: temperature converter

To practice using functions we are going to create a temperature converter. Let's start simple. Try to conceptualize the idea on paper first.

### Question 1

> Define a function called `celsius_to_kelvin` that, given a temperature in Celsius (e.g. `temp` as argument) converts it to Kelvin according to the conversion formula below. After defining the function apply it to a vector of temperatures.

$$K = °C + 273.15$$

```{r}
#| code-fold: true
#| eval: false

# define function name and arguments
celsius_to_kelvin <- function(temp) {
  
  # convert
  kelvin <- temp + 273.15
  
  # output
  return(kelvin)
  
}

x <- c(-15, -3, 0, 15, 27.5)
celsius_to_kelvin(x)
```

### Question 2

> Create the inverse function `kelvin_to_celsius` and apply it to another vector of temperatures. You will have to make sure that the temperature in Kelvin does not take negative values (since it is an absolute scale). In case this is not true, return `NA`.

```{r}
#| code-fold: true
#| eval: false

# define function name and arguments
kelvin_to_celsius <- function(temp) {
  
  # if negative in Kelvin, we stop and return absent
  # otherwise, we convert
  celsius <- if_else(temp < 0, NA, temp - 273.15)
  
  # Think why we haven't done it with an if (...) else (...)

  # output
  return(celsius)
  
}

y <- c(0, 250, 300, 350)
kelvin_to_celsius(y)
```

### Question 3

> Create a joint function `converter_temp` that has two arguments: temperature and a text argument that tells us if it is kelvin or celsius (and that by default the input temperature is Celsius). The function must use that string to decide in which direction it converts (check that the text argument does not have an option other than the two allowed; otherwise, return error using the stop(“error message...”) command). Apply it to the previous vectors and check that it gives the same.

```{r}
#| code-fold: true
#| eval: false

# define function name and arguments
# default, units in celsius
conversor_temp <- function(temp, units = "celsius") {
  
  # we check that units are correct
  # within the allowed values
  if (units %in% c("celsius", "kelvin")) {
    
    if (units == "celsius") {
      
      temp_out <- celsius_to_kelvin(temp) 
      
    } else {
      
      temp_out <- kelvin_to_celsius(temp)
      
    }
    
  } else {
    
    # otherwise we stop the function with an error message
    stop("Error: just 'celsius' or 'kelvin' as units")
  }
  
  # output
  return(temp_out)
  
}

# Notice that we have not used `if_else()` because the number of elements
# to evaluate in the condition must be equal to the number of elements that
# it returns, by doing it vectorially.
conversor_temp(x)
conversor_temp(y, units = "kelvin")
```


### Question 4

> Repeats the previous function but regardless of whether units are in upper or lower case.

```{r}
#| error: true
conversor_temp(y, units = "Kelvin")
```

```{r}
#| code-fold: true
#| eval: false

# define function name and arguments
# default, units in celsius
library(stringr)
conversor_temp <- function(temp, units = "celsius") {
  
  # we use str_to_lower to make everything lowercase
  if (str_to_lower(units) %in% c("celsius", "kelvin")) {
    
    if (units == "celsius") {
      
      temp_out <- celsius_to_kelvin(temp) 
      
    } else {
      
      temp_out <- kelvin_to_celsius(temp)
      
    }
    
  } else {
    
    # otherwise we stop the function with an error message
    stop("Error: just 'celsius' or 'kelvin' units")
  }
  
  # devolvemos
  return(temp_out)
  
}

conversor_temp(y, units = "Kelvin")
```

### Question 5


> Repeat all the above process creating `converter_temp2` but to convert between Celsius and Fahrenheit following the formula below

$$ºC = (ºF − 32) * \frac{5}{9}, \quad ºF = 32 + ºC * \frac{9}{5}$$

```{r}
#| code-fold: true
#| eval: false

celsius_to_fahr <- function(temp) {

  fahr <- 32 + temp * (9/5)
  return(fahr)
  
}
celsius_to_fahr(x)

fahr_to_celsius <- function(temp) {
  
  celsius <- (temp - 32) * (5/9)
  return(celsius)
  
}

z <- c(40, 60, 80, 100)
fahr_to_celsius(z)

conversor_temp2 <- function(temp, units = "celsius") {
  
  if (str_to_lower(units) %in% c("celsius", "fahr")) {
    
    if (units == "celsius") {
      
      temp_out <- celsius_to_fahr(temp) 
      
    } else {
      
      temp_out <- fahr_to_celsius(temp)
      
    }
    
  } else {
    
    stop("Error: just 'celsius' or 'fahr' units")
  }
  
  return(temp_out)
  
}

conversor_temp2(x)
conversor_temp2(z, units = "fahr")
```

### Question 6

> Finally, create the superfunction `converter_temp_total` that allows as input argument a temperature in one of the 3 units, a text indicating in which units it comes and another one indicating in which units it is to be output.  By default it converts from celsius to kelvin.

```{r}
#| code-fold: true
#| eval: false

converter_temp_total <-
  function(temp, units_input = "celsius",
           units_output = "kelvin") {
  
  if (str_to_lower(units_input) %in% c("celsius", "fahr", "kelvin") &
      str_to_lower(units_output) %in% c("celsius", "fahr", "kelvin")) {
    
    if (units_input == units_output) {
      
      return(temp)
      
    }
    
    else if (units_input == "celsius") {
      
      if (units_output == "kelvin") {
        
        temp_out <- celsius_to_kelvin(temp) 
        
      } else { 
        
        temp_out <- celsius_to_fahr(temp) 
      }
      
    } else if (units_input == "kelvin") {
      
      if (units_output== "celsius") {
        
        temp_out <- kelvin_to_celsius(temp) 
    
      } else { 
        
        temp_out <- celsius_to_fahr(kelvin_to_celsius(temp))
      }
      
    } else {
      
      if (units_output == "celsius") {
        
        temp_out <- fahr_to_celsius(temp) 
    
      } else { 
        
        temp_out <- celsius_to_kelvin(fahr_to_celsius(temp))
      }
      
    }
    
  } else {
    
    stop("Error: just 'celsius', 'kelvin' or 'fahr'")
  }
  
  return(temp_out)
  
}

converter_temp_total(x, units_input = "celsius",
                     units_output = "celsius")
converter_temp_total(y, units_input = "kelvin",
                     units_output = "kelvin")
converter_temp_total(y, units_input = "kelvin",
                     units_output = "celsius")
converter_temp_total(z, units_input = "fahr",
                     units_output = "celsius")

converter_temp_total(z, units_input = "fahr",
                     units_output = "celsius")
converter_temp_total(converter_temp_total(z, units_input = "fahr",
                                          units_output = "kelvin"),
                     units_input = "kelvin",
                     units_output = "celsius")
```

## 🐣 Case study III: Monty Hall problem 

In `R` the function `sample(x = ..., size = ...)` will be very useful: from a collection of `x` elements, it selects a random `size` number of them. For example, if we want to simulate 3 times the throw of a die we have 6 possible elements (`x = 1:6`) and we select it 3 times (`size = 3`).

```{r}
sample(x = 1:6, size = 3)
```

Since it is random, each time you run it, something different will come out.

```{r}
sample(x = 1:6, size = 3)
```

What if we want to throw it 10 times?

```{r}
#| error: true
sample(x = 1:6, size = 10)
```

Having only 6 possible elements and choosing 10, it cannot, so we have to indicate that we want a sample with replacement (as with the die, each face can be repeated when re-rolled).

```{r}
sample(x = 1:6, size = 10, replace = TRUE)
```

### Question 1

> With the above, imagine that you are in a TV contest where you are given a choice of 3 doors: in one there is a millionaire prize and in the other 2 an oreo cookie. Design the simulation study with for loops to approximate the probability that you get the prize (obviously it has to give you approx 0.333333333). Perform the experiment for 10, 50 trials, 100 trials, 500 trials, 1000 trials, 10 000 trials and 25 000 trials (hint: you need a loop within a loop). What do you observe?

```{r}
#| code-fold: true
#| eval: false
library(dplyr)

# Possible doors
doors <- c(1, 2, 3)

# Possible trials
trials <- c(10, 50, 100, 500, 1000, 10000, 25000)

# For scenario, we define the number of times in which we win prize
n_prizes <- rep(0, length(trials))

# first loop: scenarios
for (i in 1:length(trials)) {
  
  # second loop: for scenario, the number of trials
  for (j in 1:trials[i]) {
    
    prize <- sample(x = doors, size = 1)
    choice <- sample(x = doors, size = 1)

    n_prizes[i] <- if_else(choice == prize, n_prizes[i] + 1, n_prizes[i])
    
  }
  # in proportion
  n_prizes[i] <- n_prizes[i] / trials[i]
}
n_prizes
```

### Question 2

> What if, in each round, one of the non-winning doors that you have not chosen was opened for you, would you change doors or would you stay? Simulate both cases and find out which is the correct strategy (this problem is known as the Monty Hall problem and even appears in movies such as 21 Black Jack).


```{r}
#| code-fold: true
#| eval: false

doors <- c(1, 2, 3)
trials <- c(10, 50, 100, 500, 1000, 10000, 25000)
n_prizes_nochange <- n_prizes_change <- rep(0, length(trials))

for (i in 1:length(trials)) {
  for (j in 1:trials[i]) {
    
    init_choice <- sample(x = doors, size = 1)
    
    prize <- sample(x = doors, size = 1)
    
    open_door <-
      doors[doors != init_choice & doors != prize]
    
    if (length(open_door) > 1) {
      
      open_door <- sample(x = open_door, size = 1)
    }
      
    n_prizes_nochange[i] <-
      if_else(init_choice == prize, n_prizes_nochange[i] + 1,
              n_prizes_nochange[i])

    changed_door <- doors[doors != init_choice & doors != open_door]
    n_prizes_change[i] <-
      if_else(changed_door == prize, n_prizes_change[i] + 1,
              n_prizes_change[i])
    
  }
  n_prizes_nochange[i] <- n_prizes_nochange[i] / trials[i]
  n_prizes_change[i] <- n_prizes_change[i] / trials[i]
}
n_prizes_nochange
n_prizes_change
```



